---
title: "Habitat constraint"
author: "Lorenzo Ciannelli"
date: "3/6/2020"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir="xxxx")
knitr::opts_chunk$set(echo = T, message=FALSE, warning=FALSE, fig.align = "left")
```

```{r load libraries, include=FALSE}
getwd()
library(lattice)
library(nlme)
library(mgcv)
library(plotfunctions)
library(ggplot2)
```

##Habitat constraint analyses

The file called `Constraint_Analysis_4.csv` is the result of the constraint analyses run on each species. The last four columns of the data frame are critical. They are: 
- Rsq = the deviance explained by the geographic model
- D_MSE = the decrease in MSE as a result of including geographic coordinates in the model
- HE  = habitat extent that contains 75% of the total estimated abundnace. Habitat is consiodered to be that occuppied by ALL stages.       
- C = constraint, derived as D_MSE * (1- HE)  
Also, relative ontogeny_age = ln(1+A1)/ln(1+Amax), relative ontogeny_size = ln(1+S1)/ln(1+Smax) and liner.ontogeny = S1/smax, where S = size and A = age. 

```{r import data}
constr<-read.csv(file = 'Constraint_Analysis_4.csv',header=TRUE, check.names=TRUE)
head(constr)
```

```{r visualize data by species}
xyplot(C~liner.ontogeny|Species,data=constr)
```



**GAMM**

Try with random smooth term, with gam. Include all case studies and stages regardless of sample size.

```{r}
constr$Species<-factor(constr$Species)
gam5<-gam(C~s(liner.ontogeny,k=3)+s(Species,bs='re'),data=constr) 
summary(gam5)
AIC(gam5)
```

Plot results

```{r plot gam}
#quartz(width=7,height=5)
#par(mai=c(1,1.5,0.5,1))
plot(gam5,shade=T,select=1,res=T,pch=16,shade.col='steelblue4',col='darkorange',xlab='Ontogenetic index',ylab='Habitat constraint',cex.lab=1.5,cex.axis=1.5)
#axis(1)
#axis(2,at=c(at.i),labels=c(lb.i))
text(0.6,0.15,'P = 0.001',lwd=2,cex=1.5)
#dev.copy(jpeg,'Constraint.jpg',height=5,width=7,res=200,units='in')
#dev.off()
```


Same random gam model, but removing species/stages that have < 400 positive observations available for modeling. These include the following species and stages: 
```{r}
constr$sp_st<-paste0(constr$Species,constr$stage)
```

`r constr$sp_st[constr$sample.size<400]`. 

```{r}
gam7<-gam(C~s(liner.ontogeny,k=3)+s(Species,bs='re'),data=constr[constr$sample.size>400,]) 
summary(gam7)
AIC(gam7)
plot(gam7,shade=T,res=T,pch=16,pages=1)
gam.vcomp(gam7)
```

Remove also YFS larvae because we cannot sample their entire habitat

```{r}
gam8<-gam(C~s(liner.ontogeny,k=3)+s(Species,bs='re'),data=constr[constr$sample.size>400&constr$sp_st!="YFSlarvae",]) 
summary(gam8)
AIC(gam8)
plot(gam8,shade=T,res=T,pch=16,pages=1)
gam.vcomp(gam8)# gives the variance components for the smooth and random terms
unique(predict.gam(gam8,type="terms")[,2])# gives random terms for each species in the order shown below
unique(constr[constr$sample.size>400&constr$sp_st!="YFSlarvae","Species"])
```

Best option! But note how the variance associated with the random term is the dominant component here. 


Make a figure out of gam8.
```{r plot gam, fig.height=7,fig.width=5}
#quartz(width=5,height=7)
par(mai=c(1,1.5,0.5,1))
plot(gam8,shade=T,select=1,res=T,pch=16,shade.col='steelblue4',col='darkorange',xlab='Ontogenetic index',ylab='Habitat constraint',cex.lab=1.5,cex.axis=1.5)
text(0.6,0.15,'P < 0.001',lwd=2,cex=1.5)
#dev.copy(jpeg,'All_Constraint.jpg',height=5,width=7,res=200,units='in')
#dev.off()
```

Same as above but in bw for publication.
```{r plot gam, fig.height=7,fig.width=5}
#quartz(width=5,height=7)
#par(mai=c(1,1.5,0.5,1))
plot(gam8,shade=T,select=1,res=T,pch=16,shade.col='grey',col='black',xlab='Ontogenetic index',ylab='Habitat constraint',cex.lab=1.5,cex.axis=1.5)
text(0.6,0.15,'P < 0.001',lwd=2,cex=1.5)
#dev.copy(jpeg,'All_Constraint_BW.jpg',height=5,width=7,res=200,units='in')
#dev.off()
```


**Species GAM**

Let us first try a GAM model where species is included as a fixed effect on  both the intercept and slope. This is what I am using for the paper.

```{r}
subset<-constr[constr$sample.size>400&constr$sp_st!="YFSlarvae",]
gam9<-gam(C~factor(Species)+s(liner.ontogeny,by=factor(Species),k=3)-1,data=subset)
summary(gam9)
```

All the species-specific intercepts are significantly different from 0. Not all the smooth terms are significantly different from 0. Note how with the fixed effect formulation we get handy estimates of smooth term edf and elevation, which provide an average scaling for the overall level of habitat associations (highest in YFS for example). Let us try to plot results from this model.

```{r, fig.height=7,fig.width=10}
#quartz(height=7,width=10)
#par(mfrow=c(2,5),mai=c(0.7,0.7,0.2,0.1))
species<-as.character(unique(constr$Species))
edf.species<-summary(gam9)$edf
p.species<-anova(gam9)$s.table[,4]
aa<-sort(species)[order(p.species,decreasing=T)]
for(i in 1:length(aa)){
	newdata<-data.frame('liner.ontogeny'=seq(0,1,length=100),'Species'=rep(aa[i],100))
	pred<-predict(gam9,newdata=newdata,se.fit=T)
subset<-constr[constr$Species==aa[i]&constr$sample.size>400&constr$sp_st!="YFSlarvae",]
plot(newdata$liner.ontogeny,pred[[1]],type="l",ylim=c(0,0.6),main=aa[i],xlim=c(0,1),xlab='Ontogenetic index',ylab='Habitat constraint',cex.lab=1.3,cex.axis=1.3)
lines(newdata$liner.ontogeny,pred[[1]]+1.96*pred[[2]],lty=2)
lines(newdata$liner.ontogeny,pred[[1]]-1.96*pred[[2]],lty=2)
points(subset$liner.ontogeny,subset$C,pch=16,col='black')
if(sort(p.species,decreasing=T)[i]>0.001){
text(0.78,0.57,round(sort(p.species,decreasing=T)[i],3))}else{
text(0.78,0.57,"<0.001")}
}
#dev.copy(jpeg,'Single_Species_constraint_Fixed.jpg',height=7,width=10,res=100,units='in')
#dev.off()
```

Now let us make a figure of species intercepts, as a barplot with 95% CI
```{r}
#quartz(width=5,height=7)
intercept<-coefficients(gam9)[1:10]
int.se<-summary(gam9)$se[1:10]
ord<-order(intercept)
data.int<-data.frame("mean"=intercept[ord],"se"=int.se[ord],"name"=substr(names(intercept),16,30)[ord])
p<-ggplot(data.int) +
    geom_bar( aes(x=reorder(name,mean), y=mean), stat="identity", 
              fill="skyblue", alpha=0.7) +
    geom_errorbar( aes(x=name, ymin=mean-1.96*se, ymax=mean+1.96*se), width=0.4, colour="orange", alpha=0.9, size=1.3)
p + labs(x = "", y = "Model intercept")+ theme(axis.text.x = element_text(angle = 45, hjust = 1))
#dev.copy(jpeg,'Intercepts.jpg',height=7,width=5,res=200,units='in')
#dev.off()

```



